---
title: "CRESCYNT-Day2-Tidying"
author: "Catie Alves"
date: "March 13, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Practicing Tidying Data
## We need to first load the packages we're going to use
## We're using Dplyr first

```{r, loading.packages}
library(tidyverse)

```
### loading gapminder dataset
```{r, reading.in.data}
gapminder <-readr::read_csv('https://raw.githubusercontent.com/jules32/2017-11-30-MBARI/gh-pages/data/gapminder.csv')
```
### examining the dataset
```{r, head.tail}

head(gapminder, 2) #shows first 2 rows of dataset
tail(gapminder) #shows last rows of dataset

```
### looking at structure of data
```{r, structure}

str(gapminder) #can also get this if you look at the data in the global environment by clicking the drop down arrow
```
### dimensions of dataset
```{r, dimensions}
dim(gapminder) # gives vector of dimensions - # rows and # columns
ncol(gapminder) # number of columns
nrow(gapminder)# number of rows
```

```{r, unique}
unique(gapminder$country)
unique(gapminder$continent)
unique(gapminder[,"continent"])
```

```{r, summary}
summary(gapminder)
summary(gapminder$lifeExp)
```
### installing skimr package
```{r, skimr}
# install.packages("skimr")
library(skimr)
skim(gapminder)
```
### filtering with dplyr
```{r, filter}
dplyr::filter(gapminder, country == "Switzerland")
```
### playing with the data
```{r, exploring.data}
# we want to know which country has the 23 yr old life expectancy

dplyr::filter(gapminder, lifeExp <30) #Afghanistan and Rwanda are lowest :(

# let's look at more recent life expectancies
dplyr::filter(gapminder, country == "Afghanistan", year == max(year))
```
### selecting and deselecting the data
```{r, select}
dplyr::select(gapminder, year, lifeExp) #select a few columns of interest
```

```{r, deselect}
dplyr::select(gapminder, -continent) #drops column you don't want. Since you didn't save it, you didn't override the gapminder df
```

```{r, select.filter}
gapminder_no_cont <- dplyr::select(gapminder, -continent)
dplyr::filter(gapminder_no_cont, year == 2007)
```
### learning how to use the pipe operator
```{r, pipe operator}
dplyr::select(gapminder, -continent) %>% 
  dplyr::filter(year == 2007) 
#does same thing as above. the pipe puts the result from the first step into the second step.
```
### learning mutate function to make new columns
```{r, mutate}
# mutate allows you to compute new columns
gapminder %>% 
  mutate(gdp = gdpPercap *pop) %>% # added gdp column at the end of the df
  arrange(desc(gdp)) #shows highest; in descending order
```
### practicing using the pipe
```{r, practice}
# finding the max gdpPercap of Egypt
gapminder %>% 
  select(-continent, -lifeExp) %>% 
  filter(country == "Egypt") %>%
  mutate(gdp = pop * gdpPercap) %>% 
  mutate(max_gdp = max(gdp))

```
### group_by function
```{r, group_by}
# calculating max gdp for each country in dataset
gapminder %>% 
  group_by(country) %>% 
  mutate(gdp = pop * gdpPercap, max_gdp = max(gdp)) %>%
  ungroup()
  
```
### combining group_by and summarize functions
```{r, summarize.groupby}
gapminder %>% 
  group_by(country) %>% 
  mutate(gdp = pop * gdpPercap) %>% 
  summarize (max_gdp = max(gdp)) %>% 
  ungroup()

# basically summarize only kept the columns that were grouped_by or summarized: country and max_gdp

# Notice that the year is missing -- this basically summarized data for a single value (country) 
```
### learning how to join datasets by reading in another one (CO2) and combining it with the gap data
```{r, joining.datasets}
# reading in new data
co2 <- readr::read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/co2.csv")
```

```{r, head.summary}
head(co2)
summary(co2)
```

```{r, filtering}
# create new variable that is only 2007 data
gap_2007 <- gapminder %>% 
  dplyr::filter(year == 2007)

gap_2007

# checking that it pulled all the countries
unique(gap_2007$country)

length(unique(gap_2007$country)) #it did

```
### joining the datasets
```{r, joining}
# going to left join (gap is the main dataset, so we're left joining co2 to it)

lj <- left_join(gap_2007, co2, by= "country")
lj

rj <- right_join(gap_2007, co2, by ="country") # more NA's

# could full join to see where the full gaps in the data exist (by NA's) -- only what you do if the dataset isn't too big
```

## Now we're going to use Tidyr

```{r, tidyr}
AirPassengers 
# DEF not tidy!  What do we do?

# We want to work towards a long format where each column is a variable and each row is an observation 

# We're actually not going to tidy this df but I want to try
new.air <- tidyr::gather(as.data.frame(AirPassengers), "Month", 1:12)

new.air
```
### tidying another dataframe
```{r, tidying}
## wide format
gap_wide <- readr::read_csv('https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/gapminder_wide.csv')

## yesterday's format (intermediate)
gapminder <- readr::read_csv('https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/gapminder.csv')


```

## looking at the data
```{r, examining.data}
head(gap_wide)
str(gap_wide)

gap_long <-  gap_wide %>% 
  gather(key = obstype_year,
         value = obs_values)

str(gap_long)
head(gap_long)
tail(gap_long)
# we've reshaped our df but the format isn't what we wanted
```
```{r, gathering}
gap_wide %>% 
  gather(observation, observ_value, -continent, -country) 
# this df is better, but the observation column is weird -- no clear unit, there's a year in here; there's a what value and a when value in this column
```
 
### separating a column
```{r, separating}
gap_better <- gap_wide %>% 
  gather(observation, observ_value, -continent, -country) %>% 
  separate(observation, into = c("obs_type", "year"), sep = "_", convert = TRUE)

head(gap_better)

unique(gap_better$obs_type) # which observation types we have here 
```

### continuing to tidy data
```{r, spreading}
gap_better %>% 
  spread(key = obs_type, 
         value = observ_value)
# key = what your call names (or columns) are going to be 
# value = what are the values in the columns 
```

```{r}

```

