---
title: "Tidy_Coral_Analysis"
author: "Catie Alves"
date: "March 13, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Tidying coral data -- putting what we learned to practice!

```{r}
## load libraries
library(tidyverse)
library(janitor) #install.packages('janitor')
library(skimr) # install.packages('skimr')
library(stringr) # added when we needed it for benthic data
#install.packages("stringi")
#library(stringi)

## data filepaths/urls ----

## benthic data
benthic_url <- 'https://www.nodc.noaa.gov/archive/arc0054/0104255/1.1/data/0-data/cd08/100308OaAla03m.CSV'

## buoy data
buoy_url <- 'http://www.ndbc.noaa.gov/view_text_file.php?filename=mokh1h2010.txt.gz&dir=data/historical/stdmet/'
```

## Reading in the data from the benthic url
```{r}
benthic_raw <- read_csv(benthic_url)
head(benthic_raw)
names(benthic_raw)

```

```{r}
benthic <- benthic_raw %>% 
  janitor::clean_names() # removes capital letters and spaces!

names(benthic)
```

## cleaning benthic df

```{r}
benthic <-benthic %>% 
  select(id_name, point, x, y, id_date)

head(benthic)
```
## want to get rid of # symbols on the sides of our id_date column
```{r}
benthic <- benthic %>% 
  mutate(date = stringr::str_remove_all(id_date, "#"))

head(benthic)

# you can also do this to replace the # with blanks like below:
benthic <-  benthic %>% 
  mutate(date = stringr::str_replace_all(id_date, "#", ""))

head(benthic)
```
## Now we're gonna make a plot
```{r, benthic.plot}
ggplot(benthic, aes(date, fill=id_name))+
  geom_bar()
```

## Reading in the buoy data
```{r, buoy.data}
buoy <- read_csv(buoy_url)
head(buoy) 

# looks bad, let's try something else

buoy <- read_table2(buoy_url)
head(buoy)

# muuuuch better!

# but notice: there are 2 header rows! The second row is the units

buoy_final <- buoy
```

## We're going to fix the headers
```{r}
## 1. overwrite column names
names(buoy) <- str_c(names(buoy), # current header
                  buoy %>% filter(`#YY` == "#yr"),# row1; don't say buoy[1,] 
                  sep = "_")  ## separate by `_`

## inspect
names(buoy) ## Looks a lot better
```
## cleaning up a bit more b/c these column names are weird
```{r}
## 2. clean up a bit more to get rid of the `#`s and the `/`s. 
names(buoy) <- str_replace_all(names(buoy), "#", "")  # replace `#` with nothing 
names(buoy) <- str_replace_all(names(buoy), "/", "_") # replace `/` with `_`

## inspect to make sure it worked
names(buoy) #better!
```
## now we want to remove the first row with pseudo column names
```{r}
## 3. remove redundant row with units
buoy <- buoy %>%
  filter(YY_yr != "#yr") 

# keeps all rows that do not have "#yr"; that's because that was the label for the first row and first column! 

head(buoy) #worked!
```

## Plotting!
```{r}
# looking at sea surface temp
ggplot(buoy, aes(WTMP_degC)) +
  geom_bar()
```
## rotating tick labels
```{r}
ggplot(buoy, aes(WTMP_degC)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 90))
```
## Now we want to join the buoy data with the benthic data!
```{r}
## before we join, we want to unite the date columns of the buoy data so we can join on the date column

buoy <-  buoy %>%
  unite(date, c(YY_yr, MM_mo, DD_dy), sep ="-")

head(buoy)
```
## JOINING THE DATASETS
```{r}
bb_join <- benthic %>%
  left_join(buoy, by = "date")

head(bb_join)

## let's select a few columns and inspect:
bb_join %>%
  select(id_name, x, y, date, hh_hr, mm_mn, WTMP_degC) %>%
  head() 

## problem! bb_join is effed up
```
## we want to get mean temp per day of buoy data

1. We want to calculate mean WTMP_degC for hh_hr, mm_mn, across same date
2. To begin, we convert the characters to integers
3. Remove the 999.0 with NA
4. In the mean calculation, ignore the NA's
```{r, my.troubleshooting}

## converting WTMP to numeric
buoy$WTMP_degC <- as.numeric(buoy$WTMP_degC)

head(buoy)

## selecting only columns of interest
small_buoy <- buoy %>%
  select(date, hh_hr, mm_mn, WTMP_degC)

head(small_buoy)

##now I want to remove the NA's


  
```

## Ross came up with a method that works! See below
```{r, Ross.Method}
buoy_daily <- buoy %>%
  mutate(temp_clean = as.numeric(gsub("999.0", NA, WTMP_degC))) %>% # first need to get rid of 999's
  group_by(date) %>% # then group by date
  summarise(mean_temp = mean(temp_clean, na.rm=T)) # then get mean temp

head(buoy_daily)
```
## Now try to join the datasets since we have a date column
```{r}
bb_join_final <- benthic %>% 
  left_join(buoy_daily, by = "date")

head(bb_join_final) #has the same number of rows as benthic, so we know the join worked!
View(bb_join_final)
```
